---
title: "STA 9750 Mini-Project #01: Fiscal Characteristics of Major US Public Transit Systems"
author: "Yashvi Mittal"
date: "2025-09-30"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 110
    fig-align: center
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    dev: png
    dpi: 110
---

# Introduction

Public transit systems are public services, engineering projects, and daily economic infrastructure all at once. This report uses the National Transit Database to compare ridership, service mileage, fare revenue, and operating expenses for U.S. transit systems. The central question is not just which agency is largest, but which systems convert resources into useful service most efficiently.

```{r setup}
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("tidyverse", "readxl", "lubridate", "DT", "scales", "knitr")
missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs)

library(tidyverse)
library(readxl)
library(lubridate)
library(DT)
library(scales)
library(knitr)

dir.create("data/mp01", recursive = TRUE, showWarnings = FALSE)

download_if_missing <- function(url, dest) {
  if (file.exists(dest) && file.info(dest)$size < 10000) {
    unlink(dest)
  }

  if (!file.exists(dest)) {
    tryCatch(
      download.file(url, destfile = dest, mode = "wb", quiet = FALSE, method = "libcurl"),
      error = function(e) {
        stop(
          "Could not download ", url, ". The NTD site sometimes blocks scripted downloads; ",
          "download it in a browser and save it as ", dest, ". Original error: ",
          conditionMessage(e),
          call. = FALSE
        )
      }
    )
  }
  invisible(dest)
}
```

# Data Acquisition

The assignment provides three NTD sources: 2022 fare revenue, 2022 operating expenses, and the monthly ridership workbook. The code downloads each source only when it is not already present locally.

```{r download-data}
fare_file <- download_if_missing(
  "https://www.transit.dot.gov/sites/fta.dot.gov/files/2024-04/2022%20Fare%20Revenue.xlsx",
  "data/mp01/2022_fare_revenue.xlsx"
)

expense_file <- download_if_missing(
  "https://data.transportation.gov/api/views/dkxx-zjd6/rows.csv?date=20231102&accessType=DOWNLOAD&bom=true&format=true",
  "data/mp01/2022_expenses.csv"
)

ridership_file <- download_if_missing(
  "https://www.transit.dot.gov/sites/fta.dot.gov/files/2024-09/July%202024%20Complete%20Monthly%20Ridership%20%28with%20adjustments%20and%20estimates%29_240903.xlsx",
  "data/mp01/ridership.xlsx"
)
```

```{r prepare-financials}
FARES <- read_xlsx(fare_file) |>
  select(
    -`State/Parent NTD ID`, -`Reporter Type`, -`Reporting Module`,
    -TOS, -`Passenger Paid Fares`, -`Organization Paid Fares`
  ) |>
  filter(`Expense Type` == "Funds Earned During Period") |>
  select(-`Expense Type`) |>
  group_by(`NTD ID`, `Agency Name`, Mode) |>
  summarize(`Total Fares` = sum(`Total Fares`, na.rm = TRUE), .groups = "drop")

EXPENSES <- read_csv(expense_file, show_col_types = FALSE) |>
  select(`NTD ID`, Agency, Total, Mode) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(Expenses = Total) |>
  group_by(`NTD ID`, Mode) |>
  summarize(Expenses = sum(Expenses, na.rm = TRUE), .groups = "drop")

FINANCIALS <- inner_join(FARES, EXPENSES, join_by(`NTD ID`, Mode))
```

```{r prepare-usage}
TRIPS <- read_xlsx(ridership_file, sheet = "UPT") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, -`Reporter Type`, -`Mode/Type of Service Status`, -`UACE CD`, -TOS) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), names_to = "month", values_to = "UPT") |>
  drop_na() |>
  mutate(month = my(month))

MILES <- read_xlsx(ridership_file, sheet = "VRM") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, -`Reporter Type`, -`Mode/Type of Service Status`, -`UACE CD`, -TOS) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), names_to = "month", values_to = "VRM") |>
  drop_na() |>
  group_by(`NTD ID`, Agency, `UZA Name`, Mode, `3 Mode`, month) |>
  summarize(VRM = sum(VRM, na.rm = TRUE), .groups = "drop") |>
  mutate(month = my(month))

USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`))
```

# Cleaning

## Task 1: Creating Syntactic Names

```{r rename-columns}
USAGE <- USAGE |>
  rename(metro_area = `UZA Name`)
```

## Task 2: Recoding Transit Modes

```{r recode-mode}
USAGE |>
  distinct(Mode) |>
  arrange(Mode) |>
  kable(col.names = "Original NTD mode codes")

recode_ntd_mode <- function(x) {
  case_when(
    x == "AG" ~ "Automated Guideway",
    x == "AR" ~ "Alaska Railroad",
    x == "CB" ~ "Commuter Bus",
    x == "CC" ~ "Cable Car",
    x == "CR" ~ "Commuter Rail",
    x == "DR" ~ "Demand Response",
    x == "DT" ~ "Demand Response Taxi",
    x == "FB" ~ "Ferryboat",
    x == "HR" ~ "Heavy Rail",
    x == "IP" ~ "Inclined Plane",
    x == "LR" ~ "Light Rail",
    x == "MB" ~ "Bus",
    x == "MG" ~ "Monorail/Automated Guideway",
    x == "PB" ~ "Publico",
    x == "RB" ~ "Bus Rapid Transit",
    x == "SR" ~ "Streetcar",
    x == "TB" ~ "Trolleybus",
    x == "VP" ~ "Vanpool",
    x == "YR" ~ "Hybrid Rail",
    TRUE      ~ "Unknown"
  )
}

USAGE <- USAGE |> mutate(Mode = recode_ntd_mode(Mode))

# Apply the same recoding to FINANCIALS so Mode joins line up downstream.
FINANCIALS <- FINANCIALS |> mutate(Mode = recode_ntd_mode(Mode))
```

```{r sample-table}
USAGE |>
  select(-`NTD ID`, -`3 Mode`) |>
  rename(unlinked_passenger_trips = UPT, vehicle_revenue_miles = VRM) |>
  sample_n(min(1000, nrow(USAGE))) |>
  mutate(month = as.character(month)) |>
  datatable(options = list(pageLength = 10, scrollX = TRUE))
```

# Instructor Questions

## Task 3: Core Ridership Questions

```{r task-3-answers}
agency_most_vrm <- USAGE |>
  group_by(Agency) |>
  summarize(total_vrm = sum(VRM, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_vrm, n = 1, with_ties = FALSE)

mode_most_vrm <- USAGE |>
  group_by(Mode) |>
  summarize(total_vrm = sum(VRM, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_vrm, n = 1, with_ties = FALSE)

nyc_subway_may_2024 <- USAGE |>
  filter(
    str_detect(Agency, regex("Metropolitan Transportation Authority|MTA New York City Transit", ignore_case = TRUE)),
    Mode == "Heavy Rail",
    month == ymd("2024-05-01")
  ) |>
  summarize(trips = sum(UPT, na.rm = TRUE))

nyc_subway_april_change <- USAGE |>
  filter(
    str_detect(Agency, regex("Metropolitan Transportation Authority|MTA New York City Transit", ignore_case = TRUE)),
    Mode == "Heavy Rail",
    month %in% ymd(c("2019-04-01", "2020-04-01"))
  ) |>
  group_by(month) |>
  summarize(trips = sum(UPT, na.rm = TRUE), .groups = "drop") |>
  arrange(month) |>
  mutate(previous_trips = lag(trips), percent_change = (trips - previous_trips) / previous_trips)

mode_longest_avg_trip <- USAGE |>
  filter(month == ymd("2024-05-01"), UPT > 0, VRM > 0) |>
  group_by(Mode) |>
  summarize(
    avg_trip_miles = sum(VRM, na.rm = TRUE) / sum(UPT, na.rm = TRUE),
    .groups = "drop"
  ) |>
  slice_max(avg_trip_miles, n = 1, with_ties = FALSE)

bind_rows(
  tibble(question = "Agency with most total VRM", answer = agency_most_vrm$Agency, value = comma(agency_most_vrm$total_vrm)),
  tibble(question = "Mode with most total VRM", answer = mode_most_vrm$Mode, value = comma(mode_most_vrm$total_vrm)),
  tibble(question = "NYC subway trips in May 2024", answer = "NYC Subway / Heavy Rail", value = comma(nyc_subway_may_2024$trips)),
  tibble(question = "Mode with longest average trip, May 2024", answer = mode_longest_avg_trip$Mode, value = paste0(round(mode_longest_avg_trip$avg_trip_miles, 2), " mi/trip")),
  tibble(question = "NYC subway ridership change, April 2019 to April 2020", answer = "April 2020 relative to April 2019", value = percent(last(na.omit(nyc_subway_april_change$percent_change)), accuracy = 0.1))
) |>
  kable()
```

The first two answers describe the scale of service production. The NYC subway questions show the pandemic interruption: the April 2020 collapse is a structural shock, not a normal year-over-year fluctuation. I was genuinely surprised by how dominant a single mode and a single agency are — before running this I expected more spread across cities.

To put the NYC pandemic shock in context, the chart below traces NYC subway monthly ridership from 2018 to the most recent month in the data:

```{r nyc-subway-timeline}
nyc_subway_timeline <- USAGE |>
  filter(
    str_detect(Agency, regex("Metropolitan Transportation Authority|MTA New York City Transit", ignore_case = TRUE)),
    Mode == "Heavy Rail",
    year(month) >= 2018
  ) |>
  group_by(month) |>
  summarize(trips = sum(UPT, na.rm = TRUE), .groups = "drop")

nyc_subway_timeline |>
  ggplot(aes(x = month, y = trips)) +
  geom_line(color = "#2c3e50", linewidth = 0.9) +
  geom_point(color = "#e74c3c", size = 0.8) +
  geom_vline(xintercept = as.numeric(ymd("2020-03-01")), linetype = "dashed", color = "grey40") +
  annotate("text", x = ymd("2020-04-01"), y = max(nyc_subway_timeline$trips) * 0.85,
           label = "COVID-19\nshutdown", color = "grey20", hjust = 0, size = 3.2) +
  scale_y_continuous(labels = comma) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "NYC Subway Monthly Ridership, 2018-Present",
    subtitle = "The pandemic dip is large enough that no normal recession compares",
    x = NULL, y = "Unlinked passenger trips"
  ) +
  theme_minimal(base_size = 12)
```

## Task 4: Three Additional Transit Facts

```{r extra-facts}
latest_month <- max(USAGE$month, na.rm = TRUE)

top_metros_latest <- USAGE |>
  filter(month == latest_month) |>
  group_by(metro_area) |>
  summarize(trips = sum(UPT, na.rm = TRUE), .groups = "drop") |>
  slice_max(trips, n = 10)

rail_bus_mix <- USAGE |>
  filter(year(month) == 2024) |>
  mutate(category = if_else(Mode %in% c("Heavy Rail", "Commuter Rail", "Light Rail", "Streetcar"), "Rail", "Non-rail")) |>
  group_by(metro_area, category) |>
  summarize(trips = sum(UPT, na.rm = TRUE), .groups = "drop") |>
  group_by(metro_area) |>
  mutate(total_trips = sum(trips), rail_share = if_else(category == "Rail", trips / total_trips, NA_real_)) |>
  ungroup() |>
  filter(total_trips >= 10000000, category == "Rail") |>
  slice_max(rail_share, n = 10)

service_intensity <- USAGE |>
  filter(year(month) == 2024) |>
  group_by(Agency, Mode) |>
  summarize(UPT = sum(UPT, na.rm = TRUE), VRM = sum(VRM, na.rm = TRUE), .groups = "drop") |>
  filter(VRM > 0, UPT >= 1000000) |>
  mutate(trips_per_vrm = UPT / VRM) |>
  slice_max(trips_per_vrm, n = 10)

top_metros_latest |>
  ggplot(aes(x = reorder(metro_area, trips), y = trips)) +
  geom_col(fill = "#2c7fb8", width = 0.75) +
  geom_text(aes(label = comma(trips, accuracy = 1)), hjust = -0.1, size = 3, color = "grey30") +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Largest Transit Markets in the Latest Available Month",
    subtitle = paste("Data through", format(latest_month, "%B %Y")),
    x = NULL,
    y = "Unlinked passenger trips"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

### Fact 1: The largest transit markets dominate national ridership

```{r fact-table-1}
top_metros_latest |>
  mutate(trips = comma(trips)) |>
  datatable(
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE,
    colnames = c("Metro Area", "Unlinked Passenger Trips")
  )
```

### Fact 2: Some large metros are much more rail-oriented

```{r fact-table-2}
rail_bus_mix |>
  select(metro_area, total_trips, rail_share) |>
  mutate(
    total_trips = comma(total_trips),
    rail_share = percent(rail_share, accuracy = 0.1)
  ) |>
  datatable(
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE,
    colnames = c("Metro Area", "Total Trips (2024)", "Rail Share")
  )
```

### Fact 3: Trips per vehicle mile identify intensively used service

```{r fact-table-3}
service_intensity |>
  select(Agency, Mode, UPT, VRM, trips_per_vrm) |>
  mutate(
    UPT = comma(UPT),
    VRM = comma(VRM),
    trips_per_vrm = round(trips_per_vrm, 2)
  ) |>
  datatable(
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE,
    colnames = c("Agency", "Mode", "Unlinked Passenger Trips", "Vehicle Revenue Miles", "Trips per VRM")
  )
```

# Annual 2022 Usage and Finance

## Task 5: Table Summarization

```{r annual-summary}
USAGE_2022_ANNUAL <- USAGE |>
  filter(year(month) == 2022) |>
  group_by(`NTD ID`, Agency, metro_area, Mode) |>
  summarize(UPT = sum(UPT, na.rm = TRUE), VRM = sum(VRM, na.rm = TRUE), .groups = "drop")

USAGE_AND_FINANCIALS <- left_join(
  USAGE_2022_ANNUAL,
  FINANCIALS,
  join_by(`NTD ID`, Mode)
) |>
  drop_na() |>
  mutate(
    farebox_recovery = `Total Fares` / Expenses,
    expense_per_upt = Expenses / UPT,
    fare_per_upt = `Total Fares` / UPT,
    expense_per_vrm = Expenses / VRM,
    fare_per_vrm = `Total Fares` / VRM
  )

USAGE_2022_ANNUAL |>
  slice_max(UPT, n = 10) |>
  kable(digits = 2)
```

## Task 6: Farebox Recovery Among Major Systems

I define a major system as any agency-mode pair with at least 400,000 annual unlinked passenger trips. This threshold removes tiny systems whose ratios can swing wildly from one accounting entry. I tried a few different cutoffs — 100k, 400k, 1M — and 400k felt like the right balance between including real transit systems and excluding the outlier demand-response operations that have almost no riders but very high per-trip costs.

```{r farebox-comparisons}
major_systems <- USAGE_AND_FINANCIALS |>
  filter(UPT >= 400000, VRM > 0, Expenses > 0)

summary_questions <- bind_rows(
  major_systems |> slice_max(UPT, n = 1, with_ties = FALSE) |> mutate(metric = "Most UPT in 2022"),
  major_systems |> slice_max(farebox_recovery, n = 1, with_ties = FALSE) |> mutate(metric = "Highest farebox recovery"),
  major_systems |> slice_min(expense_per_upt, n = 1, with_ties = FALSE) |> mutate(metric = "Lowest expenses per UPT"),
  major_systems |> slice_max(fare_per_upt, n = 1, with_ties = FALSE) |> mutate(metric = "Highest total fares per UPT"),
  major_systems |> slice_min(expense_per_vrm, n = 1, with_ties = FALSE) |> mutate(metric = "Lowest expenses per VRM"),
  major_systems |> slice_max(fare_per_vrm, n = 1, with_ties = FALSE) |> mutate(metric = "Highest total fares per VRM")
) |>
  select(metric, Agency, Mode, metro_area, UPT, VRM, `Total Fares`, Expenses, farebox_recovery, expense_per_upt, fare_per_upt, expense_per_vrm, fare_per_vrm)

summary_questions |>
  mutate(across(c(UPT, VRM, `Total Fares`, Expenses), comma)) |>
  mutate(across(contains("per_"), dollar)) |>
  mutate(farebox_recovery = percent(farebox_recovery, accuracy = 0.1)) |>
  datatable(
    options = list(pageLength = 6, dom = "t", scrollX = TRUE),
    rownames = FALSE,
    caption = "Major 2022 transit systems — winner on each efficiency metric"
  )
```

```{r efficiency-plot}
major_systems |>
  filter(UPT >= quantile(UPT, 0.75, na.rm = TRUE)) |>
  ggplot(aes(x = expense_per_upt, y = farebox_recovery, size = UPT, color = Mode)) +
  geom_point(alpha = 0.75) +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = percent) +
  scale_size_continuous(labels = comma) +
  labs(
    title = "Farebox Recovery Rewards Both Ridership and Fare Revenue",
    subtitle = "Major 2022 agency-mode systems",
    x = "Operating expense per unlinked passenger trip",
    y = "Farebox recovery ratio",
    size = "UPT",
    color = "Mode"
  ) +
  theme_minimal()
```

# Conclusion

The most efficient system depends on how you define efficiency, and I found that no single agency wins on every metric. Farebox recovery rewards systems that collect high fares, while expenses per trip rewards systems that move a lot of people cheaply — and those two goals can pull in opposite directions. I give the most weight to expenses per passenger trip because it directly measures how cheaply a system provides useful rides, regardless of fare policy. That said, the farebox recovery numbers were genuinely surprising to me: some of the systems I expected to be well-run had shockingly low recovery, which made me reconsider what "efficient" actually means in a public-service context. The most convincing candidates for overall efficiency are systems that sit in the top half on multiple metrics rather than dominating just one.
