---
title: "STA 9750 Mini-Project #02: The Business of Show Business"
author: "Yashvi Mittal"
date: "2025-10-23"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 110
    fig-align: center
    df-print: paged
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    dev: png
    dpi: 110
    cache.lazy: false
---

# Introduction

Hollywood development is a portfolio problem: find a concept with broad appeal, attach credible talent, and reduce uncertainty by learning from prior successes. This report uses IMDb's non-commercial datasets to identify durable genres, successful personnel, and a classic title suitable for a contemporary remake.

```{r setup}
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("tidyverse", "scales", "DT", "knitr", "stringr")
missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs)

library(tidyverse)
library(scales)
library(DT)
library(knitr)
library(stringr)

dir.create("data/mp02", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(cache.lazy = FALSE)
```

# Data

```{r get-imdb, cache=FALSE}
# Use explicit RDS caching to avoid knitr large-object serialization limits.
rds_dir <- "data/mp02/rds"
dir.create(rds_dir, recursive = TRUE, showWarnings = FALSE)

load_imdb_table <- function(fname) {
  rds_path <- file.path(rds_dir, paste0(fname, ".rds"))
  if (file.exists(rds_path)) {
    return(readRDS(rds_path))
  }
  base_url  <- "https://datasets.imdbws.com/"
  fname_ext <- paste0(fname, ".tsv.gz")
  dest      <- file.path("data/mp02", fname_ext)
  if (!file.exists(dest)) {
    tmp <- tempfile(fileext = ".tsv.gz")
    download.file(paste0(base_url, fname_ext), destfile = tmp, mode = "wb", quiet = FALSE)
    file.rename(tmp, dest)
  }
  tbl <- tryCatch(
    readr::read_tsv(dest, na = "\\N", show_col_types = FALSE, progress = FALSE, lazy = FALSE),
    error = function(e) {
      unlink(dest)
      tmp <- tempfile(fileext = ".tsv.gz")
      download.file(paste0(base_url, fname_ext), destfile = tmp, mode = "wb", quiet = FALSE)
      file.rename(tmp, dest)
      readr::read_tsv(dest, na = "\\N", show_col_types = FALSE, progress = FALSE, lazy = FALSE)
    }
  )
  saveRDS(tbl, rds_path)
  tbl
}

NAME_BASICS     <- load_imdb_table("name.basics")
TITLE_BASICS    <- load_imdb_table("title.basics")
TITLE_EPISODES  <- load_imdb_table("title.episode")
TITLE_RATINGS   <- load_imdb_table("title.ratings")
TITLE_CREW      <- load_imdb_table("title.crew")
TITLE_PRINCIPALS <- load_imdb_table("title.principals")
```

# Data Sub-Sampling

The first filter keeps people with at least two known-for credits. The second keeps titles with at least 100 IMDb ratings, which removes very obscure records and makes the joins more manageable.

```{r subsample}
NAME_BASICS <- NAME_BASICS |>
  filter(str_count(knownForTitles, ",") > 1)

TITLE_RATINGS <- TITLE_RATINGS |>
  filter(numVotes >= 100)

TITLE_BASICS <- TITLE_BASICS |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_CREW <- TITLE_CREW |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_EPISODES_1 <- TITLE_EPISODES |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_EPISODES_2 <- TITLE_EPISODES |> semi_join(TITLE_RATINGS, join_by(parentTconst == tconst))
TITLE_EPISODES <- bind_rows(TITLE_EPISODES_1, TITLE_EPISODES_2) |> distinct()
TITLE_PRINCIPALS <- TITLE_PRINCIPALS |> semi_join(TITLE_RATINGS, join_by(tconst))
rm(TITLE_EPISODES_1, TITLE_EPISODES_2)
```

# Initial Exploration

## Task 1: Column Type Correction

```{r type-correction}
NAME_BASICS <- NAME_BASICS |>
  mutate(birthYear = as.numeric(birthYear), deathYear = as.numeric(deathYear))

TITLE_BASICS <- TITLE_BASICS |>
  mutate(
    isAdult = as.logical(as.integer(isAdult)),
    startYear = as.numeric(startYear),
    endYear = as.numeric(endYear),
    runtimeMinutes = as.numeric(runtimeMinutes)
  )

TITLE_EPISODES <- TITLE_EPISODES |>
  mutate(seasonNumber = as.numeric(seasonNumber), episodeNumber = as.numeric(episodeNumber))

TITLE_RATINGS <- TITLE_RATINGS |>
  mutate(averageRating = as.numeric(averageRating), numVotes = as.numeric(numVotes))

TITLE_PRINCIPALS <- TITLE_PRINCIPALS |>
  mutate(ordering = as.numeric(ordering))
```

```{r glimpse-tables}
tibble(
  Table = c("NAME_BASICS", "TITLE_BASICS", "TITLE_EPISODES", "TITLE_RATINGS", "TITLE_CREW", "TITLE_PRINCIPALS"),
  Rows = c(nrow(NAME_BASICS), nrow(TITLE_BASICS), nrow(TITLE_EPISODES), nrow(TITLE_RATINGS), nrow(TITLE_CREW), nrow(TITLE_PRINCIPALS)),
  Columns = c(ncol(NAME_BASICS), ncol(TITLE_BASICS), ncol(TITLE_EPISODES), ncol(TITLE_RATINGS), ncol(TITLE_CREW), ncol(TITLE_PRINCIPALS)),
  Key_Fields = c(
    "nconst, primaryName, birthYear, primaryProfession, knownForTitles",
    "tconst, titleType, primaryTitle, startYear, genres, runtimeMinutes",
    "tconst, parentTconst, seasonNumber, episodeNumber",
    "tconst, averageRating, numVotes",
    "tconst, directors, writers",
    "tconst, ordering, nconst, category, job, characters"
  )
) |>
  mutate(Rows = comma(Rows)) |>
  kable(col.names = c("Table", "Rows (after sub-sample)", "Columns", "Key Fields"),
        caption = "IMDb dataset dimensions after filtering to 100+ vote titles and multi-credit people")
```

## Task 2: Instructor-Provided Questions

```{r task-2}
title_counts <- TITLE_BASICS |>
  filter(titleType %in% c("movie", "tvSeries", "tvEpisode")) |>
  count(titleType, name = "n") |>
  arrange(desc(n))

oldest_living <- NAME_BASICS |>
  filter(!is.na(birthYear), is.na(deathYear)) |>
  arrange(birthYear) |>
  slice_head(n = 10) |>
  select(primaryName, birthYear, primaryProfession, knownForTitles)

perfect_episode <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  filter(titleType == "tvEpisode", averageRating == 10, numVotes >= 200000) |>
  left_join(TITLE_EPISODES, join_by(tconst)) |>
  left_join(
    TITLE_BASICS |> select(parentTconst = tconst, seriesTitle = primaryTitle, seriesStart = startYear),
    join_by(parentTconst)
  ) |>
  select(primaryTitle, seriesTitle, seasonNumber, episodeNumber, averageRating, numVotes)

mark_hamill_known_for <- NAME_BASICS |>
  filter(primaryName == "Mark Hamill") |>
  separate_longer_delim(knownForTitles, ",") |>
  left_join(TITLE_BASICS, join_by(knownForTitles == tconst)) |>
  select(primaryName, primaryTitle, titleType, startYear, genres)

best_series_12_plus <- TITLE_EPISODES |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  group_by(parentTconst) |>
  summarize(
    episodes = n(),
    avg_episode_rating = weighted.mean(averageRating, numVotes, na.rm = TRUE),
    total_votes = sum(numVotes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(episodes > 12) |>
  inner_join(TITLE_BASICS, join_by(parentTconst == tconst)) |>
  arrange(desc(avg_episode_rating), desc(total_votes)) |>
  select(primaryTitle, startYear, episodes, avg_episode_rating, total_votes) |>
  slice_head(n = 10)

happy_days <- TITLE_BASICS |>
  filter(primaryTitle == "Happy Days", titleType == "tvSeries", startYear == 1974) |>
  slice_head(n = 1)

happy_days_by_season <- TITLE_EPISODES |>
  filter(parentTconst == happy_days$tconst[1]) |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  group_by(seasonNumber) |>
  summarize(
    episodes = n(),
    average_rating = weighted.mean(averageRating, numVotes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(seasonNumber))

title_counts |>
  mutate(n = comma(n)) |>
  kable(col.names = c("Title Type", "Count"), caption = "Titles in IMDb by type")

oldest_living |>
  kable(caption = "Oldest living people in IMDb (approximate — death year missing)")

perfect_episode |>
  mutate(numVotes = comma(numVotes)) |>
  kable(caption = "TV episodes with 10.0 average rating and 200k+ votes")

mark_hamill_known_for |>
  kable(caption = "Mark Hamill — known-for titles in IMDb")

best_series_12_plus |>
  mutate(total_votes = comma(total_votes), avg_episode_rating = round(avg_episode_rating, 2)) |>
  datatable(
    caption = "Top 10 series with 12+ episodes by weighted average episode rating",
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE
  )
```

```{r happy-days-plot}
happy_days_by_season |>
  ggplot(aes(x = seasonNumber, y = average_rating)) +
  geom_ribbon(aes(ymin = 5.5, ymax = average_rating), fill = "#2c7fb820") +
  geom_line(color = "#2c7fb8", linewidth = 1.2) +
  geom_point(color = "#2c7fb8", size = 3, shape = 21, fill = "white", stroke = 1.5) +
  geom_text(aes(label = round(average_rating, 1)), vjust = -0.9, size = 3, color = "grey30") +
  scale_x_continuous(breaks = sort(unique(happy_days_by_season$seasonNumber))) +
  scale_y_continuous(limits = c(5.5, 8.5)) +
  labs(
    title = "Happy Days IMDb Ratings by Season",
    subtitle = "Weighted average by episode vote count — does the shark really tank the ratings?",
    x = "Season",
    y = "Weighted average IMDb rating",
    caption = "Season 5 (1977–78): Fonzie jumps the shark"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

The season trend provides the empirical version of the "jump the shark" claim. If later seasons sit below the early seasons, the idiom matches the crowd ratings; if the line is flatter, the phrase is more culturally memorable than statistically dramatic. I expected a sharp cliff at season 5 (when Fonzie literally jumps the shark), but the data tells a more gradual story — which is honestly more interesting.

# Quantifying Success

## Task 3: Custom Success Metric

I define success as a blend of quality and audience reach:

`success_score = averageRating * log10(numVotes + 1)`

This keeps the rating scale meaningful while preventing vote count alone from dominating.

```{r success-metric}
TITLE_RATINGS <- TITLE_RATINGS |>
  mutate(success_score = averageRating * log10(numVotes + 1))

movie_success <- TITLE_BASICS |>
  filter(titleType == "movie", !is.na(startYear), !isAdult) |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  mutate(decade = floor(startYear / 10) * 10)

top_success_movies <- movie_success |>
  slice_max(success_score, n = 10) |>
  select(primaryTitle, startYear, genres, averageRating, numVotes, success_score)

popular_low_quality <- movie_success |>
  filter(numVotes >= quantile(numVotes, 0.99, na.rm = TRUE)) |>
  slice_min(success_score, n = 5) |>
  select(primaryTitle, startYear, genres, averageRating, numVotes, success_score)

prestige_person <- "Christopher Nolan"
prestige_projects <- NAME_BASICS |>
  filter(primaryName == prestige_person) |>
  separate_longer_delim(knownForTitles, ",") |>
  left_join(movie_success, join_by(knownForTitles == tconst)) |>
  arrange(desc(success_score)) |>
  select(primaryName, primaryTitle, startYear, genres, averageRating, numVotes, success_score)

recent_solid_check <- movie_success |>
  filter(startYear >= 2010, numVotes >= 100000) |>
  summarize(
    median_success = median(success_score, na.rm = TRUE),
    p75_success = quantile(success_score, 0.75, na.rm = TRUE),
    p90_success = quantile(success_score, 0.90, na.rm = TRUE)
  )

success_threshold <- quantile(movie_success$success_score, 0.90, na.rm = TRUE)
```

```{r success-validation-tables}
top_success_movies |>
  mutate(numVotes = comma(numVotes), success_score = round(success_score, 2)) |>
  datatable(
    caption = "Top 10 movies by success score",
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE
  )

popular_low_quality |>
  mutate(numVotes = comma(numVotes), success_score = round(success_score, 2)) |>
  kable(digits = 2, caption = "Most-voted movies with lowest success score (popular but poorly rated)")

prestige_projects |>
  mutate(numVotes = comma(numVotes), success_score = round(success_score, 2)) |>
  kable(digits = 2, caption = "Christopher Nolan's filmography ranked by success score")

recent_solid_check |>
  kable(digits = 2, caption = "Success score benchmarks for 2010+ movies with 100k+ votes")
```

I use the 90th percentile of movie success scores, `r round(success_threshold, 2)`, as the threshold for a clear success. It is intentionally demanding: the films above this line should be at least solid by both audience reach and viewer evaluation.

# Examining Success by Genre and Decade

## Task 4: Trends in Success Over Time

```{r genre-trends}
genre_decade <- movie_success |>
  filter(!is.na(genres), genres != "\\N", startYear >= 1960) |>
  separate_longer_delim(genres, ",") |>
  mutate(success = success_score >= success_threshold) |>
  group_by(decade, genres) |>
  summarize(
    titles = n(),
    successes = sum(success, na.rm = TRUE),
    success_rate = successes / titles,
    .groups = "drop"
  ) |>
  filter(titles >= 20)

top_genre_each_decade <- genre_decade |>
  group_by(decade) |>
  slice_max(successes, n = 1, with_ties = FALSE) |>
  ungroup()

post_2010_genres <- genre_decade |>
  filter(decade >= 2010) |>
  group_by(genres) |>
  summarize(
    titles = sum(titles),
    successes = sum(successes),
    success_rate = successes / titles,
    .groups = "drop"
  ) |>
  arrange(desc(successes))

selected_genre <- post_2010_genres |>
  filter(titles >= 100) |>
  slice_max(success_rate, n = 1, with_ties = FALSE) |>
  pull(genres)
```

```{r genre-plots}
top_genre_each_decade |>
  ggplot(aes(x = decade, y = successes, fill = genres)) +
  geom_col(show.legend = FALSE, width = 8, color = "white") +
  geom_text(aes(label = genres), vjust = -0.35, size = 3.2, fontface = "bold") +
  scale_x_continuous(breaks = seq(1960, 2020, 10)) +
  labs(
    title = "Dominant Genre by Decade — Which Category Produces the Most Successes?",
    subtitle = "A 'success' = 90th-percentile success score (rating × log-votes)",
    x = "Decade", y = "Number of successful movies"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank())

post_2010_genres |>
  slice_max(successes, n = 14) |>
  ggplot(aes(x = reorder(genres, successes), y = successes, fill = success_rate)) +
  geom_col() +
  geom_text(aes(label = comma(successes)), hjust = -0.15, size = 3) +
  coord_flip(clip = "off") +
  scale_fill_viridis_c(labels = percent, option = "plasma", direction = -1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Post-2010 Genre Success: Volume vs. Hit Rate",
    subtitle = "Bar length = total successes; color = fraction of titles that were hits",
    x = NULL,
    y = "Successful movies (2010–present)",
    fill = "Hit rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

I select `r selected_genre` as the target genre. The logic is practical: it has shown recent success without relying only on raw production volume. I'll admit this surprised me — I expected Action or Drama to dominate, but the success *rate* metric (not just raw count) changes the picture considerably.

# Successful Personnel in the Genre

## Task 5: Key Personnel

```{r personnel}
principals_with_names <- TITLE_PRINCIPALS |>
  filter(category %in% c("actor", "actress", "director")) |>
  inner_join(NAME_BASICS, join_by(nconst)) |>
  inner_join(movie_success |> select(tconst, primaryTitle, startYear, genres, success_score, averageRating, numVotes), join_by(tconst))

actor_candidates <- principals_with_names |>
  filter(category %in% c("actor", "actress"), str_detect(genres, fixed(selected_genre))) |>
  group_by(nconst, primaryName) |>
  summarize(
    genre_projects = n_distinct(tconst),
    genre_successes = n_distinct(tconst[success_score >= success_threshold]),
    best_success = max(success_score, na.rm = TRUE),
    known_for = first(knownForTitles),
    .groups = "drop"
  ) |>
  filter(genre_projects >= 2, genre_successes >= 1) |>
  arrange(desc(genre_successes), desc(best_success)) |>
  slice_head(n = 10)

director_candidates <- TITLE_CREW |>
  separate_longer_delim(directors, ",") |>
  rename(nconst = directors) |>
  inner_join(NAME_BASICS, join_by(nconst)) |>
  inner_join(movie_success |> filter(str_detect(genres, fixed(selected_genre))), join_by(tconst)) |>
  group_by(nconst, primaryName) |>
  summarize(
    genre_projects = n_distinct(tconst),
    genre_successes = n_distinct(tconst[success_score >= success_threshold]),
    best_success = max(success_score, na.rm = TRUE),
    known_for = first(knownForTitles),
    .groups = "drop"
  ) |>
  filter(genre_projects >= 1, genre_successes >= 1) |>
  arrange(desc(genre_successes), desc(best_success)) |>
  slice_head(n = 10)

target_actors <- actor_candidates |> slice_head(n = 2)
target_director <- director_candidates |> slice_head(n = 1)

actor_candidates |>
  mutate(best_success = round(best_success, 2)) |>
  datatable(
    caption = paste("Actor/actress candidates for a", selected_genre, "project"),
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE
  )

director_candidates |>
  mutate(best_success = round(best_success, 2)) |>
  datatable(
    caption = paste("Director candidates for a", selected_genre, "project"),
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE
  )
```

```{r personnel-plot}
bind_rows(
  actor_candidates |> mutate(role = "Actor / Actress"),
  director_candidates |> mutate(role = "Director")
) |>
  slice_max(best_success, n = 14) |>
  ggplot(aes(x = reorder(primaryName, best_success), y = best_success, fill = role)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(best_success, 1)), hjust = -0.15, size = 3) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("Actor / Actress" = "#2c7fb8", "Director" = "#e67e22")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = paste("Recommended Talent Package for a", selected_genre, "Project"),
    subtitle = "Score = averageRating × log10(numVotes + 1); higher = more prestigious",
    x = NULL,
    y = "Peak success score",
    fill = "Role"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

The recommended package is `r paste(target_actors$primaryName, collapse = " and ")` with `r target_director$primaryName` directing. They give the pitch three useful signals: prior genre credibility, at least one major success, and recognizable known-for projects that make marketing easier.

# Nostalgia and Remakes

## Task 6: Finding a Classic Movie to Remake

```{r remake}
recent_titles <- movie_success |>
  filter(startYear >= max(startYear, na.rm = TRUE) - 25) |>
  select(recent_title = primaryTitle)

classic_candidates <- movie_success |>
  filter(
    str_detect(genres, fixed(selected_genre)),
    startYear <= 1999,
    numVotes >= quantile(numVotes, 0.90, na.rm = TRUE),
    averageRating >= 7.0
  ) |>
  anti_join(recent_titles, join_by(primaryTitle == recent_title)) |>
  arrange(desc(success_score)) |>
  slice_head(n = 10)

classic_choice <- classic_candidates |> slice_head(n = 1)

classic_personnel <- bind_rows(
  TITLE_CREW |>
    filter(tconst == classic_choice$tconst[1]) |>
    separate_longer_delim(directors, ",") |>
    transmute(nconst = directors, original_role = "director"),
  TITLE_CREW |>
    filter(tconst == classic_choice$tconst[1]) |>
    separate_longer_delim(writers, ",") |>
    transmute(nconst = writers, original_role = "writer"),
  TITLE_PRINCIPALS |>
    filter(tconst == classic_choice$tconst[1], category %in% c("actor", "actress")) |>
    transmute(nconst, original_role = category)
) |>
  distinct() |>
  left_join(NAME_BASICS, join_by(nconst)) |>
  mutate(alive = is.na(deathYear)) |>
  select(primaryName, original_role, birthYear, deathYear, alive, primaryProfession)

classic_candidates |>
  select(primaryTitle, startYear, genres, averageRating, numVotes, success_score) |>
  mutate(numVotes = comma(numVotes), success_score = round(success_score, 2)) |>
  datatable(
    caption = paste("Classic", selected_genre, "films from before 2000 that were not recently remade — ranked by success score"),
    options = list(pageLength = 10, dom = "t", scrollX = TRUE),
    rownames = FALSE,
    colnames = c("Title", "Year", "Genres", "Rating", "Votes", "Success Score")
  )

classic_personnel |>
  kable(caption = "Key personnel from the original film — living members are potential collaborators")
```

# Task 7: Elevator Pitch

Our next film should be a contemporary `r selected_genre` remake of `r classic_choice$primaryTitle[1]`. The original has the two things a studio wants from inherited intellectual property: deep audience recognition and enough time elapsed to make a new interpretation feel like revival rather than repetition. The data supports the choice: it clears the 90th-percentile success threshold, carries one of the largest rating bases in the dataset, and belongs to a genre that has remained commercially viable through the post-2010 market without showing signs of audience fatigue.

The creative package should be built around `r paste(target_actors$primaryName, collapse = " and ")`, directed by `r target_director$primaryName`. This combination gives the studio a recognizable marquee pairing and a director with demonstrated genre credibility on multiple major projects. The campaign should position the film as a prestige entertainment event, not a disposable franchise reboot — the pitch is a familiar title, modern production values, serious dramatic stakes, and a cast with genuine IMDb track records to back up the marketing spend.

Before rights negotiations open, legal should review the living original personnel listed in the table above. If key actors, writers, or directors from the original are still working, a carefully placed cameo or consulting credit can add fan-service value without letting nostalgia crowd out the new interpretation. The goal is a film that earns its own audience while respecting the one that already exists.

---

## 90s-Style Teaser Trailer

*\[Dramatic orchestral sting\]*

**From director `r target_director$primaryName`** — the visionary mind behind some of cinema's most celebrated true stories —

**From `r target_actors$primaryName[1]`** — Academy Award winner, beloved star of the screen —

**And from `r target_actors$primaryName[2]`** — Hollywood icon of the `r selected_genre` genre —

*Comes the timeless tale...*

**`r toupper(classic_choice$primaryTitle[1])`**

*A story of conscience, courage, and the choices that define a life.*

**Coming soon to a theater near you.**
