---
title: "STA 9750 Mini-Project #02: The Business of Show Business"
author: "Yashvi Mittal"
date: "2025-10-23"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
  cache: true
---

# Introduction

Hollywood development is a portfolio problem: find a concept with broad appeal, attach credible talent, and reduce uncertainty by learning from prior successes. This report uses IMDb's non-commercial datasets to identify durable genres, successful personnel, and a classic title suitable for a contemporary remake.

```{r setup}
options(repos = c(CRAN = "https://cloud.r-project.org"))
pkgs <- c("tidyverse", "scales", "DT", "knitr", "stringr")
missing_pkgs <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs)

library(tidyverse)
library(scales)
library(DT)
library(knitr)
library(stringr)

dir.create("data/mp02", recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(cache.lazy = FALSE)
```

# Data

```{r get-imdb}
get_imdb_file <- function(fname) {
  base_url <- "https://datasets.imdbws.com/"
  fname_ext <- paste0(fname, ".tsv.gz")
  dest <- file.path("data/mp02", fname_ext)
  download_imdb <- function() {
    tmp <- tempfile(fileext = ".tsv.gz")
    download.file(paste0(base_url, fname_ext), destfile = tmp, mode = "wb", quiet = FALSE)
    file.rename(tmp, dest)
  }

  if (!file.exists(dest)) {
    download_imdb()
  }

  tryCatch(
    readr::read_tsv(dest, na = "\\N", show_col_types = FALSE, progress = FALSE, lazy = FALSE),
    error = function(e) {
      unlink(dest)
      download_imdb()
      readr::read_tsv(dest, na = "\\N", show_col_types = FALSE, progress = FALSE, lazy = FALSE)
    }
  )
}

NAME_BASICS <- get_imdb_file("name.basics")
TITLE_BASICS <- get_imdb_file("title.basics")
TITLE_EPISODES <- get_imdb_file("title.episode")
TITLE_RATINGS <- get_imdb_file("title.ratings")
TITLE_CREW <- get_imdb_file("title.crew")
TITLE_PRINCIPALS <- get_imdb_file("title.principals")
```

# Data Sub-Sampling

The first filter keeps people with at least two known-for credits. The second keeps titles with at least 100 IMDb ratings, which removes very obscure records and makes the joins more manageable.

```{r subsample}
NAME_BASICS <- NAME_BASICS |>
  filter(str_count(knownForTitles, ",") > 1)

TITLE_RATINGS <- TITLE_RATINGS |>
  filter(numVotes >= 100)

TITLE_BASICS <- TITLE_BASICS |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_CREW <- TITLE_CREW |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_EPISODES_1 <- TITLE_EPISODES |> semi_join(TITLE_RATINGS, join_by(tconst))
TITLE_EPISODES_2 <- TITLE_EPISODES |> semi_join(TITLE_RATINGS, join_by(parentTconst == tconst))
TITLE_EPISODES <- bind_rows(TITLE_EPISODES_1, TITLE_EPISODES_2) |> distinct()
TITLE_PRINCIPALS <- TITLE_PRINCIPALS |> semi_join(TITLE_RATINGS, join_by(tconst))
rm(TITLE_EPISODES_1, TITLE_EPISODES_2)
```

# Initial Exploration

## Task 1: Column Type Correction

```{r type-correction}
NAME_BASICS <- NAME_BASICS |>
  mutate(birthYear = as.numeric(birthYear), deathYear = as.numeric(deathYear))

TITLE_BASICS <- TITLE_BASICS |>
  mutate(
    isAdult = as.logical(as.integer(isAdult)),
    startYear = as.numeric(startYear),
    endYear = as.numeric(endYear),
    runtimeMinutes = as.numeric(runtimeMinutes)
  )

TITLE_EPISODES <- TITLE_EPISODES |>
  mutate(seasonNumber = as.numeric(seasonNumber), episodeNumber = as.numeric(episodeNumber))

TITLE_RATINGS <- TITLE_RATINGS |>
  mutate(averageRating = as.numeric(averageRating), numVotes = as.numeric(numVotes))

TITLE_PRINCIPALS <- TITLE_PRINCIPALS |>
  mutate(ordering = as.numeric(ordering))
```

```{r glimpse-tables}
list(
  NAME_BASICS = NAME_BASICS,
  TITLE_BASICS = TITLE_BASICS,
  TITLE_EPISODES = TITLE_EPISODES,
  TITLE_RATINGS = TITLE_RATINGS,
  TITLE_CREW = TITLE_CREW,
  TITLE_PRINCIPALS = TITLE_PRINCIPALS
) |>
  imap(function(tbl, nm) {
    cat("\n\n### ", nm, "\n\n", sep = "")
    print(glimpse(tbl))
  })
```

## Task 2: Instructor-Provided Questions

```{r task-2}
title_counts <- TITLE_BASICS |>
  filter(titleType %in% c("movie", "tvSeries", "tvEpisode")) |>
  count(titleType, name = "n") |>
  arrange(desc(n))

oldest_living <- NAME_BASICS |>
  filter(!is.na(birthYear), is.na(deathYear)) |>
  arrange(birthYear) |>
  slice_head(n = 10) |>
  select(primaryName, birthYear, primaryProfession, knownForTitles)

perfect_episode <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  filter(titleType == "tvEpisode", averageRating == 10, numVotes >= 200000) |>
  left_join(TITLE_EPISODES, join_by(tconst)) |>
  left_join(
    TITLE_BASICS |> select(parentTconst = tconst, seriesTitle = primaryTitle, seriesStart = startYear),
    join_by(parentTconst)
  ) |>
  select(primaryTitle, seriesTitle, seasonNumber, episodeNumber, averageRating, numVotes)

mark_hamill_known_for <- NAME_BASICS |>
  filter(primaryName == "Mark Hamill") |>
  separate_longer_delim(knownForTitles, ",") |>
  left_join(TITLE_BASICS, join_by(knownForTitles == tconst)) |>
  select(primaryName, primaryTitle, titleType, startYear, genres)

best_series_12_plus <- TITLE_EPISODES |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  group_by(parentTconst) |>
  summarize(
    episodes = n(),
    avg_episode_rating = weighted.mean(averageRating, numVotes, na.rm = TRUE),
    total_votes = sum(numVotes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(episodes > 12) |>
  inner_join(TITLE_BASICS, join_by(parentTconst == tconst)) |>
  arrange(desc(avg_episode_rating), desc(total_votes)) |>
  select(primaryTitle, startYear, episodes, avg_episode_rating, total_votes) |>
  slice_head(n = 10)

happy_days <- TITLE_BASICS |>
  filter(primaryTitle == "Happy Days", titleType == "tvSeries", startYear == 1974) |>
  slice_head(n = 1)

happy_days_by_season <- TITLE_EPISODES |>
  filter(parentTconst == happy_days$tconst[1]) |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  group_by(seasonNumber) |>
  summarize(
    episodes = n(),
    average_rating = weighted.mean(averageRating, numVotes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(seasonNumber))

title_counts |> kable()
oldest_living |> kable()
perfect_episode |> kable()
mark_hamill_known_for |> kable()
best_series_12_plus |> kable(digits = 2)
```

```{r happy-days-plot}
happy_days_by_season |>
  ggplot(aes(x = seasonNumber, y = average_rating)) +
  geom_line(color = "#2c7fb8", linewidth = 1) +
  geom_point(color = "#2c7fb8", size = 2) +
  scale_x_continuous(breaks = sort(unique(happy_days_by_season$seasonNumber))) +
  labs(
    title = "Happy Days Ratings by Season",
    subtitle = "Weighted by episode vote count",
    x = "Season",
    y = "Average IMDb rating"
  ) +
  theme_minimal()
```

The season trend provides the empirical version of the "jump the shark" claim. If later seasons sit below the early seasons, the idiom matches the crowd ratings; if the line is flatter, the phrase is more culturally memorable than statistically dramatic.

# Quantifying Success

## Task 3: Custom Success Metric

I define success as a blend of quality and audience reach:

`success_score = averageRating * log10(numVotes + 1)`

This keeps the rating scale meaningful while preventing vote count alone from dominating.

```{r success-metric}
TITLE_RATINGS <- TITLE_RATINGS |>
  mutate(success_score = averageRating * log10(numVotes + 1))

movie_success <- TITLE_BASICS |>
  filter(titleType == "movie", !is.na(startYear), !isAdult) |>
  inner_join(TITLE_RATINGS, join_by(tconst)) |>
  mutate(decade = floor(startYear / 10) * 10)

top_success_movies <- movie_success |>
  slice_max(success_score, n = 10) |>
  select(primaryTitle, startYear, genres, averageRating, numVotes, success_score)

popular_low_quality <- movie_success |>
  filter(numVotes >= quantile(numVotes, 0.99, na.rm = TRUE)) |>
  slice_min(success_score, n = 5) |>
  select(primaryTitle, startYear, genres, averageRating, numVotes, success_score)

prestige_person <- "Christopher Nolan"
prestige_projects <- NAME_BASICS |>
  filter(primaryName == prestige_person) |>
  separate_longer_delim(knownForTitles, ",") |>
  left_join(movie_success, join_by(knownForTitles == tconst)) |>
  arrange(desc(success_score)) |>
  select(primaryName, primaryTitle, startYear, genres, averageRating, numVotes, success_score)

recent_solid_check <- movie_success |>
  filter(startYear >= 2010, numVotes >= 100000) |>
  summarize(
    median_success = median(success_score, na.rm = TRUE),
    p75_success = quantile(success_score, 0.75, na.rm = TRUE),
    p90_success = quantile(success_score, 0.90, na.rm = TRUE)
  )

success_threshold <- quantile(movie_success$success_score, 0.90, na.rm = TRUE)
```

```{r success-validation-tables}
top_success_movies |> kable(digits = 2)
popular_low_quality |> kable(digits = 2)
prestige_projects |> kable(digits = 2)
recent_solid_check |> kable(digits = 2)
```

I use the 90th percentile of movie success scores, `r round(success_threshold, 2)`, as the threshold for a clear success. It is intentionally demanding: the films above this line should be at least solid by both audience reach and viewer evaluation.

# Examining Success by Genre and Decade

## Task 4: Trends in Success Over Time

```{r genre-trends}
genre_decade <- movie_success |>
  filter(!is.na(genres), genres != "\\N", startYear >= 1960) |>
  separate_longer_delim(genres, ",") |>
  mutate(success = success_score >= success_threshold) |>
  group_by(decade, genres) |>
  summarize(
    titles = n(),
    successes = sum(success, na.rm = TRUE),
    success_rate = successes / titles,
    .groups = "drop"
  ) |>
  filter(titles >= 20)

top_genre_each_decade <- genre_decade |>
  group_by(decade) |>
  slice_max(successes, n = 1, with_ties = FALSE) |>
  ungroup()

post_2010_genres <- genre_decade |>
  filter(decade >= 2010) |>
  group_by(genres) |>
  summarize(
    titles = sum(titles),
    successes = sum(successes),
    success_rate = successes / titles,
    .groups = "drop"
  ) |>
  arrange(desc(successes))

selected_genre <- post_2010_genres |>
  filter(titles >= 100) |>
  slice_max(success_rate, n = 1, with_ties = FALSE) |>
  pull(genres)
```

```{r genre-plots}
top_genre_each_decade |>
  ggplot(aes(x = decade, y = successes, fill = genres)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = genres), vjust = -0.2, size = 3) +
  labs(title = "Most Successful Genre by Decade", x = "Decade", y = "Number of successful movies") +
  theme_minimal()

post_2010_genres |>
  slice_max(successes, n = 12) |>
  ggplot(aes(x = reorder(genres, successes), y = successes, fill = success_rate)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(labels = percent) +
  labs(
    title = "Post-2010 Genre Success",
    subtitle = "Counts and rates can tell different stories",
    x = NULL,
    y = "Successful movies",
    fill = "Success rate"
  ) +
  theme_minimal()
```

I select `r selected_genre` as the target genre. The logic is practical: it has shown recent success without relying only on raw production volume.

# Successful Personnel in the Genre

## Task 5: Key Personnel

```{r personnel}
principals_with_names <- TITLE_PRINCIPALS |>
  filter(category %in% c("actor", "actress", "director")) |>
  inner_join(NAME_BASICS, join_by(nconst)) |>
  inner_join(movie_success |> select(tconst, primaryTitle, startYear, genres, success_score, averageRating, numVotes), join_by(tconst))

actor_candidates <- principals_with_names |>
  filter(category %in% c("actor", "actress"), str_detect(genres, fixed(selected_genre))) |>
  group_by(nconst, primaryName) |>
  summarize(
    genre_projects = n_distinct(tconst),
    genre_successes = n_distinct(tconst[success_score >= success_threshold]),
    best_success = max(success_score, na.rm = TRUE),
    known_for = first(knownForTitles),
    .groups = "drop"
  ) |>
  filter(genre_projects >= 2, genre_successes >= 1) |>
  arrange(desc(genre_successes), desc(best_success)) |>
  slice_head(n = 10)

director_candidates <- TITLE_CREW |>
  separate_longer_delim(directors, ",") |>
  rename(nconst = directors) |>
  inner_join(NAME_BASICS, join_by(nconst)) |>
  inner_join(movie_success |> filter(str_detect(genres, fixed(selected_genre))), join_by(tconst)) |>
  group_by(nconst, primaryName) |>
  summarize(
    genre_projects = n_distinct(tconst),
    genre_successes = n_distinct(tconst[success_score >= success_threshold]),
    best_success = max(success_score, na.rm = TRUE),
    known_for = first(knownForTitles),
    .groups = "drop"
  ) |>
  filter(genre_projects >= 1, genre_successes >= 1) |>
  arrange(desc(genre_successes), desc(best_success)) |>
  slice_head(n = 10)

target_actors <- actor_candidates |> slice_head(n = 2)
target_director <- director_candidates |> slice_head(n = 1)

actor_candidates |> kable(digits = 2)
director_candidates |> kable(digits = 2)
```

```{r personnel-plot}
bind_rows(
  actor_candidates |> mutate(role = "Actor"),
  director_candidates |> mutate(role = "Director")
) |>
  slice_max(best_success, n = 12) |>
  ggplot(aes(x = reorder(primaryName, best_success), y = best_success, fill = role)) +
  geom_col() +
  coord_flip() +
  labs(
    title = paste("Potential Talent for a", selected_genre, "Project"),
    x = NULL,
    y = "Best success score",
    fill = "Role"
  ) +
  theme_minimal()
```

The recommended package is `r paste(target_actors$primaryName, collapse = " and ")` with `r target_director$primaryName` directing. They give the pitch three useful signals: prior genre credibility, at least one major success, and recognizable known-for projects that make marketing easier.

# Nostalgia and Remakes

## Task 6: Finding a Classic Movie to Remake

```{r remake}
recent_titles <- movie_success |>
  filter(startYear >= max(startYear, na.rm = TRUE) - 25) |>
  select(recent_title = primaryTitle)

classic_candidates <- movie_success |>
  filter(
    str_detect(genres, fixed(selected_genre)),
    startYear <= 1999,
    numVotes >= quantile(numVotes, 0.90, na.rm = TRUE),
    averageRating >= 7.0
  ) |>
  anti_join(recent_titles, join_by(primaryTitle == recent_title)) |>
  arrange(desc(success_score)) |>
  slice_head(n = 10)

classic_choice <- classic_candidates |> slice_head(n = 1)

classic_personnel <- bind_rows(
  TITLE_CREW |>
    filter(tconst == classic_choice$tconst[1]) |>
    separate_longer_delim(directors, ",") |>
    transmute(nconst = directors, original_role = "director"),
  TITLE_CREW |>
    filter(tconst == classic_choice$tconst[1]) |>
    separate_longer_delim(writers, ",") |>
    transmute(nconst = writers, original_role = "writer"),
  TITLE_PRINCIPALS |>
    filter(tconst == classic_choice$tconst[1], category %in% c("actor", "actress")) |>
    transmute(nconst, original_role = category)
) |>
  distinct() |>
  left_join(NAME_BASICS, join_by(nconst)) |>
  mutate(alive = is.na(deathYear)) |>
  select(primaryName, original_role, birthYear, deathYear, alive, primaryProfession)

classic_candidates |> select(primaryTitle, startYear, genres, averageRating, numVotes, success_score) |> kable(digits = 2)
classic_personnel |> kable()
```

# Task 7: Elevator Pitch

Our next film should be a contemporary `r selected_genre` remake of `r classic_choice$primaryTitle[1]`. The original has the two things a studio wants from inherited intellectual property: deep audience recognition and enough age to make a new interpretation feel like revival rather than repetition. The data supports the choice because it clears the success threshold, has a large rating base, and belongs to a genre that has remained commercially viable in the post-2010 market.

The creative package should be built around `r paste(target_actors$primaryName, collapse = " and ")`, directed by `r target_director$primaryName`. This combination gives the studio a recognizable marquee pairing and a director with demonstrated genre credibility. The campaign should sell the film as a prestige entertainment event rather than a disposable reboot: familiar title, modern stakes, and talent with proven IMDb success.

Before rights negotiations, legal should review the living original personnel listed above. If any original actors, writers, or directors are available, a carefully chosen cameo or consulting credit could add fan-service value without letting nostalgia crowd out the new film. The pitch is simple: use a proven story, attach credible genre talent, and market the project as the rare remake with both an audience and a reason to exist.
